---
title: "Transcriptomic Data Analysis in Mosquito Larvae"
subtitle: "Microplastic exposure alters immune-related gene expression in *Culex quinquefasciatus* mosquitoes and larval microbiota diversity"
author: "Ahamed TCHATAKOURA, Marie BUYSSE, Marie-Laure SETIER RIO, Olivier ROUX, Claire LOISEAU, Amandine AVILES"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: cerulean
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo      = TRUE,
  message   = FALSE,
  warning   = FALSE,
  fig.align = "center",
  fig.width  = 8,
  fig.height = 5,
  dpi       = 150
)
```

---

# 0. Loading libraries

```{r libraries}
library(DESeq2)
library(ggplot2)
library(pheatmap)
library(reshape2)
library(ashr)
library(rtracklayer)
library(ggrepel)
```

---

# 1. Global settings

## 1.1 Theme

```{r theme-pub}
themePlot <- function(base_size = 12) {
  theme_classic(base_size = base_size) +
    theme(
      plot.title         = element_text(face = "bold", size = base_size + 2,
                                        hjust = 0.5, margin = margin(b = 8)),
      plot.subtitle      = element_text(size = base_size, hjust = 0.5,
                                        color = "grey40", margin = margin(b = 6)),
      axis.title         = element_text(face = "bold", size = base_size),
      axis.text          = element_text(size = base_size - 1, color = "black"),
      axis.line          = element_line(linewidth = 0.5, color = "black"),
      axis.ticks         = element_line(linewidth = 0.4, color = "black"),
      legend.title       = element_text(face = "bold", size = base_size - 1),
      legend.text        = element_text(size = base_size - 2),
      legend.key.size    = unit(0.45, "cm"),
      legend.background  = element_blank(),
      legend.position    = "right",
      panel.grid.major.y = element_line(color = "grey92", linewidth = 0.3),
      panel.grid.major.x = element_blank(),
      panel.grid.minor   = element_blank(),
      plot.margin        = margin(10, 15, 10, 10)
    )
}
```

## 1.2 Colour palette (conditions)

```{r palette}
palette_conditions <- c(
  "CTRL" = "#4D4DFF",
  "60"   = "#2E86AB",
  "200"  = "#E07A5F",
  "600"  = "#3D405B"
)
```

---

# 2. Data import & preparation

```{r import}
counts_larv <- read.table("larves.txt", header = TRUE, sep = "\t",
                           row.names = 1, check.names = FALSE)
meta_larv   <- read.csv("metadataL.csv", header = TRUE, sep = ",",
                         row.names = 1, stringsAsFactors = FALSE,
                         check.names = FALSE)

# Factor with explicit level order
meta_larv$condition <- factor(meta_larv$condition,
                               levels = c("CTRL", "60", "200", "600"))

# Harmonise sample order (metadata ↔ counts)
meta_larv$replicate <- as.integer(sub(".*_(\\d+)$", "\\1", rownames(meta_larv)))
meta_larv   <- meta_larv[order(meta_larv$condition, meta_larv$replicate), , drop = FALSE]
counts_larv <- counts_larv[, rownames(meta_larv)]
stopifnot(identical(rownames(meta_larv), colnames(counts_larv)))
meta_larv$replicate <- NULL
```

---

# 3. Quality control

## 3.1 Library sizes

The library size plot below includes all samples. `controlL_rep3` shows a markedly
lower sequencing depth than the other replicates and is removed before downstream
analyses (section 3.2).

```{r libsize, fig.cap="Total counts per sample, coloured by condition. The low-depth outlier (controlL_rep3) is visible and excluded in the next step."}
libsize <- colSums(counts_larv)
df_libsize <- data.frame(
  sample       = factor(names(libsize), levels = rownames(meta_larv)),
  total_counts = libsize,
  condition    = meta_larv$condition
)

ggplot(df_libsize, aes(x = sample, y = total_counts / 1e6, fill = condition)) +
  geom_col(width = 0.75, color = "white", linewidth = 0.2) +
  scale_fill_manual(values = palette_conditions, name = "Condition") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
  labs(
    title = "Sequencing depth per sample",
    x     = NULL,
    y     = "Total counts (× 10⁶)"
  ) +
  themePlot() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## 3.2 Outlier removal

```{r outlier-removal}
# controlL_rep3 excluded: library size too low (visible in 3.1)
bad         <- "controlL_rep3"
counts_larv <- counts_larv[, colnames(counts_larv) != bad]
meta_larv   <- meta_larv[rownames(meta_larv) != bad, , drop = FALSE]
```

## 3.3 DESeq2 object & low-count filtering

```{r deseq-build}
dds_larv <- DESeqDataSetFromMatrix(
  countData = counts_larv,
  colData   = meta_larv,
  design    = ~ condition
)

# Keep genes with ≥ 10 counts in ≥ 2 samples
keep     <- rowSums(counts(dds_larv) >= 10) >= 2
dds_larv <- dds_larv[keep, ]

cat(sprintf("Genes before filtering : %d\n", nrow(counts_larv)))
cat(sprintf("Genes after filtering  : %d\n", nrow(dds_larv)))
```

## 3.4 VST normalisation

```{r vst}
dds_larv <- estimateSizeFactors(dds_larv)
vsd_larv <- vst(dds_larv, blind = FALSE)
```

## 3.5 VST distribution per sample

```{r vst-dist, fig.cap="VST-normalised expression distribution. Violin + boxplot per sample.", fig.height=5.5}
df_vst       <- as.data.frame(assay(vsd_larv))
df_vst$gene  <- rownames(df_vst)
dfl          <- melt(df_vst, id.vars = "gene",
                     variable.name = "sample", value.name = "vst")
dfl$condition <- meta_larv[as.character(dfl$sample), "condition"]

ggplot(dfl, aes(x = sample, y = vst, fill = condition)) +
  geom_violin(alpha = 0.6, linewidth = 0.3) +
  geom_boxplot(width = 0.12, outlier.shape = NA,
               fill = "white", linewidth = 0.4) +
  scale_fill_manual(values = palette_conditions, name = "Condition") +
  labs(
    title = "VST-normalised expression - larvae",
    x     = NULL,
    y     = "VST counts"
  ) +
  themePlot() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## 3.6 Principal Component Analysis

```{r pca, fig.cap="PCA on VST-transformed counts. Each point is one biological replicate."}
plotPCA(vsd_larv, intgroup = "condition") +
  scale_color_manual(values = palette_conditions, name = "Condition") +
  labs(title = "Principal Component Analysis - larvae") +
  themePlot()
```

---

# 4. Differential expression analysis

```{r deseq-run}
dds_larv <- DESeq(dds_larv)

res_larv_60  <- lfcShrink(dds_larv, contrast = c("condition", "60",  "CTRL"), type = "ashr")
res_larv_200 <- lfcShrink(dds_larv, contrast = c("condition", "200", "CTRL"), type = "ashr")
res_larv_600 <- lfcShrink(dds_larv, contrast = c("condition", "600", "CTRL"), type = "ashr")
```

## 4.1 MA plots

```{r ma-plots, fig.show="hold", out.width="33%", fig.cap="MA plots for each contrast (ashr LFC shrinkage). Red dots: padj < 0.05."}
plotMA(res_larv_60,  main = "MA plot - 60 MPs/mL vs CTRL",  alpha = 0.05)
plotMA(res_larv_200, main = "MA plot - 200 MPs/mL vs CTRL", alpha = 0.05)
plotMA(res_larv_600, main = "MA plot - 600 MPs/mL vs CTRL", alpha = 0.05)
```

## 4.2 Volcano plots

```{r volcano-fn}
plot_volcano <- function(res, title, n_label = 15, lfc_thresh = 1, padj_thresh = 0.05) {

  df <- subset(as.data.frame(res), !is.na(padj) & !is.na(log2FoldChange))
  df$gene <- rownames(df)

  df$regulation <- "NS"
  df$regulation[df$padj < padj_thresh & df$log2FoldChange >  lfc_thresh] <- "Up"
  df$regulation[df$padj < padj_thresh & df$log2FoldChange < -lfc_thresh] <- "Down"
  df$regulation <- factor(df$regulation, levels = c("Up", "Down", "NS"))

  n_up   <- sum(df$regulation == "Up")
  n_down <- sum(df$regulation == "Down")

  df_sig   <- df[df$regulation != "NS", ]
  df_sig   <- df_sig[order(df_sig$padj), ]
  top_up   <- head(df_sig[df_sig$regulation == "Up",   ], n_label)
  top_down <- head(df_sig[df_sig$regulation == "Down", ], n_label)
  df_label <- rbind(top_up, top_down)

  ggplot(df, aes(x = log2FoldChange, y = -log10(padj), color = regulation)) +
    geom_point(data = df[df$regulation == "NS", ], alpha = 0.4, size = 1.2) +
    geom_point(data = df[df$regulation != "NS", ], alpha = 0.8, size = 1.8) +
    geom_vline(xintercept = c(-lfc_thresh, lfc_thresh),
               linetype = "dashed", color = "grey40", linewidth = 0.5) +
    geom_hline(yintercept = -log10(padj_thresh),
               linetype = "dashed", color = "grey40", linewidth = 0.5) +
    geom_text_repel(
      data          = df_label,
      aes(label     = gene),
      size          = 4,
      max.overlaps  = 20,
      box.padding   = 0.4,
      point.padding = 0.5,
      segment.color = "grey50",
      segment.size  = 0.3,
      show.legend   = FALSE,
      fontface      = "italic"
    ) +
    geom_point(data = df[df$regulation == "NS", ],
           color = "grey60", alpha = 0.6, size = 1.5) +
    scale_color_manual(
      values = c("Up" = "red", "Down" = "blue", "NS" = "lightgrey"),
      labels = c(
        "Up"   = paste0("Up (n=",   n_up,   ")"),
        "Down" = paste0("Down (n=", n_down, ")"),
        "NS"   = "NS"
      )
    ) +
    coord_cartesian(xlim = c(-max(abs(df$log2FoldChange)) * 1.05,
                              max(abs(df$log2FoldChange)) * 1.05)) +
    theme_classic(base_size = 13) +
    theme(
      legend.position = "right",
      legend.title    = element_blank(),
      plot.title      = element_text(face = "bold", hjust = 0.5),
      axis.line       = element_line(color = "black")
    ) +
    labs(
      title = title,
      x     = expression(log[2]~"Fold Change"),
      y     = expression(-log[10]~"(padj)")
    )
}
```

```{r volcano-60, fig.cap="Volcano plot - 60 MPs/mL vs CTRL. Top 15 up- and down-regulated genes labelled."}
plot_volcano(res_larv_60, "60 MPs/mL vs CTRL - larvae")
```

```{r volcano-200, fig.cap="Volcano plot - 200 MPs/mL vs CTRL."}
plot_volcano(res_larv_200, "200 MPs/mL vs CTRL - larvae")
```

```{r volcano-600, fig.cap="Volcano plot - 600 MPs/mL vs CTRL."}
plot_volcano(res_larv_600, "600 MPs/mL vs CTRL - larvae")
```

---

# 5. DEG extraction

```{r degs}
extract_degs <- function(res, lfc_threshold = 1, padj_threshold = 0.05) {
  subset(res, padj < padj_threshold & abs(log2FoldChange) > lfc_threshold)
}

deg_larv_60  <- extract_degs(res_larv_60)
deg_larv_200 <- extract_degs(res_larv_200)
deg_larv_600 <- extract_degs(res_larv_600)

genes_larv_60  <- rownames(deg_larv_60)
genes_larv_200 <- rownames(deg_larv_200)
genes_larv_600 <- rownames(deg_larv_600)

cat(sprintf("DEGs - 60 vs CTRL  : %d\n", length(genes_larv_60)))
cat(sprintf("DEGs - 200 vs CTRL : %d\n", length(genes_larv_200)))
cat(sprintf("DEGs - 600 vs CTRL : %d\n", length(genes_larv_600)))
```

```{r degs-print}
print(genes_larv_60)
print(genes_larv_200)
print(genes_larv_600)
```
